---
title: "Spectral Counting Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r include = FALSE}
library(conflicted)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(scales)

library(MSstats)
library(edgeR)
library(qvalue)

library(plotly)
library(DT)
library(htmltools)

theme_set(theme_bw(base_size = 10))
```



```{r}
## EdgeR
PSMFrame <- data.frame(PSMtable)
row.names(PSMFrame) <- PSMFrame$accession
PSMFrame <- PSMFrame[, -1]
PSMFrame[is.na(PSMFrame)] <- 0

# remove rows with less than 10 counts
OKtest <- rowSums(PSMFrame) > 0
filteredPSMFrame <- PSMFrame[OKtest, ]

## sort csv2mzTab so that rows correspond to PSMFrame columns
csv2mzTabSorted <- left_join(data.frame(conditionRep = colnames(PSMFrame)),
                             csv2mzTab %>% mutate(conditionRep = make.names(conditionRep)),
                             by = "conditionRep")

## prepare design
group <- factor(csv2mzTabSorted$Condition)
y <- DGEList(counts=filteredPSMFrame, group=group)
y <- calcNormFactors(y)
design <- model.matrix(~ 0 + group)
colnames(design) <- gsub("group", "", colnames(design))

## Make sure design matrix corresponds to contrast mat
if(!all(colnames(contrast_mat) == colnames(design))){
  stop("Error in correspondance between contrast and design matrix")
}

## Estimate dispersion and perform quasi-likelihood F-tests:
y <- estimateDisp(y, design)
fit <- glmQLFit(y, design)

## Iterate through the contrasts
## Iterate through the contrasts
qlfList <- list()
qObjList <- list()
for(i in 1:nrow(contrast_mat)){
  qlfList[[i]] <- glmQLFTest(fit, contrast = contrast_mat[i,])
  qObjList[[i]] <- qvalue(qlfList[[i]]$table$PValue)
}

```


`r nrow(PSMFrame)` Proteins were detected.  `r sum(OKtest)` had any quantitative data.  `r ncol(PSMFrame)` biological samples were analyzed from `r length(unique(data$Reference))` raw files.  EdgeR was used to identify affected proteins.  

```{r}
## counts by biosample
data.frame(sample = colnames(PSMtable)[-1],
           count = colSums(PSMtable[, -1], na.rm = TRUE)) %>%
  mutate(group = sapply(strsplit(sample, "_"), '[', 1)) %>%
  ggplot(aes(x = sample, y = count, fill = group)) +
  ## color samples by group alternating light and dark grey
  scale_fill_manual(values = rep(c(grey(0.3), grey(0.6)), 
                                 ceiling(length(unique(group))/2))[1:length(unique(group))]) +
  geom_col() +
  theme(legend.position = 'none') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  coord_flip()
```

**Figure 1** Number of spectral counts per biological sample.  

---

## Summary of affected proteins

```{r}
summaryTable <- data.frame(Contrast = row.names(contrast_mat), q10 = 0,
                           q30 = 0, estimateChanged = 0)

for(i in 1:nrow(summaryTable)){
  summaryTable$q10[i] <- sum(qObjList[[i]]$qvalues < 0.1, na.rm = TRUE)
  summaryTable$q30[i] <- sum(qObjList[[i]]$qvalues < 0.3, na.rm = TRUE)
  summaryTable$estimateChanged[i] <- round((1-qObjList[[i]]$pi0)*nrow(filteredPSMFrame), 1)
}
DT::datatable(summaryTable, rownames = FALSE,
              colnames = c("Contrast", "q < 0.1", "q < 0.3", "Estimate Changed"))
```

---

## p-value histogram    
p-value histogram for the contrast `r rownames(contrast_mat)[whichContrast]`.  Horizontal line indicates the number of true null hypotheses per bin.  

```{r}
pValHistFun <- function(qlfRes){
  yintercept <- nrow(qlfRes$table) * qvalue(qlfRes$table$PValue)$pi0/20
  title <- rownames(contrast_mat)[whichContrast]
  qlfRes$table %>%
    ggplot(aes(x = PValue)) +
    geom_histogram(bins = 20) +
    geom_hline(yintercept = yintercept) +
    labs(title = title)
}

pValHistFun(qlfList[[whichContrast]])
```

---

## Volcano plot  
Volcano plot for the contrast `r rownames(contrast_mat)[whichContrast]`.  The plot is interactive: hovering will bring up protein ID information.

```{r}
## get proteins for description
protein <- proteins(MzTabRes) %>%
  mutate(description = sapply(strsplit(description, " OS"), '[', 1))

volcanoFun <- function(qlfRes, title){
  qlfRes$table %>%
    mutate(accession = row.names(qlfRes$table)) %>%
    inner_join(dplyr::select(protein, accession, description), by = "accession") %>%
    mutate(qValue = qvalue(PValue)$qvalues) %>%
    mutate(affected = qValue < 0.1) %>%
    mutate(FC = 2 ^ logFC) %>%
    mutate(negLogP = -log10(PValue)) %>%
    ggplot(aes(x = FC, y = negLogP, text = description)) +
    geom_point(data = function(x) x[!x$affected,], color = "black") +
    geom_point(data = function(x) x[x$affected,], color = "red") +
    scale_x_continuous(trans = "log2", breaks = breaks_log(n = 7, base = 2)) +
    labs(x = "Fold change", y = "-log10 p-value", title = title)
}

ggplotly(volcanoFun(qlfList[[whichContrast]], title = rownames(contrast_mat)[whichContrast]))

```




```{r}
# MAplot <- qlf$table %>%
#   mutate(intensity = rowSums(PSMFrame[OKtest, ])) %>%
#   mutate(accession = row.names(qlf$table)) %>%
#   inner_join(dplyr::select(protein, accession, description), by = "accession") %>%
#   mutate(FC = 2^logFC) %>%
#   mutate(qValues = qvalue(PValue)$qvalues) %>%
#   mutate(changed = !is.na(qValues) & qValues < 0.001) %>%
#   mutate(PValue = -log10(PValue)) %>%
#   ggplot(aes(x = intensity, y = FC, color = changed, text = description)) +
#   geom_point() +
#   scale_color_manual(values = c("black", "red")) +
#   scale_y_continuous(trans = "log2", breaks = 2^seq(-4, 8, 2)) +
#   scale_x_continuous(trans = "log10") +
#   labs(x = "Total Counts", y = "Fold-change")
# 
# ggplotly(MAplot)

```

```{r}
## Single output table
qlfList[[whichContrast]]$table %>%
  mutate(accession = row.names(.)) %>%
  inner_join(dplyr::select(protein, accession, description), by = "accession") %>%
  mutate(label = rownames(contrast_mat)[whichContrast]) %>%
  relocate(label) %>%
  relocate(description) %>%
  relocate(accession) %>%
  write.csv(paste0(make.names(rownames(contrast_mat)[whichContrast]), ".csv"))
```


```{r}
## pooled output table
if(whichContrast == length(qlfList)){
  outputTableList <- list()
  for(i in 1:length(qlfList)){
    label <- rownames(contrast_mat)[i]
    outputTableList[[i]] <- qlfList[[i]]$table %>%
      mutate(accession = row.names(.)) %>%
      inner_join(dplyr::select(protein, accession, description), by = "accession") %>%
      mutate(label = label) %>%
      relocate(label) %>%
      relocate(description) %>%
      relocate(accession) 
  }
  do.call(rbind, outputTableList) %>%
    write.csv("edgeR_output.csv")
}


```

